---
title: "Preliminary Early Onset Gastric Cancer Analysis"
author: "Madeline Dabney"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(ggplot2)
library(tidyverse)
library(skimr)
library(janitor)
library(dplyr)
library(forcats)
library(knitr)
library(tidyr)
```

```{r, echo=FALSE}
data <- read.csv('gastric_cancer_data.csv', stringsAsFactors = TRUE, na.strings = c("", "N/A", "NA"))
```

```{r, echo=FALSE, warning=FALSE}
# Clean data 
data <- data %>%
  mutate(across(where(is.factor), ~ fct_recode(
    factor(trimws(.x)), 
    "N/A" = "N/A ", 
    "N" = "N ", 
    "Y" = "Y "
  )))

# Recode race and ethnicity categories
data <- data %>%
  mutate(across(where(is.factor), ~ fct_recode(.x,
    `Male` = "1  Male", `Female` = "2 Female", `Female` = "2  Female",
    `White` = "01  White", `Black or African American` = "02  Black or African American",
    `Vietnamese` = "10  Vietnamese", `Asian Indian` = "16  Asian Indian",
    `Other Asian` = "96  Other Asian, including Asian, NOS", `Some other race` = "98  Some other race",
    `Non-Spanish` = "0  Non-Spanish; non-Hispanic", `Mexican` = "1  Mexican (includes Chicano)",
    `South or Central American` = "4  South or Central American (except Brazil)",
    `Spanish` = "6  Spanish, NOS; Hispanic, NOS; Latino, NOS",
    `In` = "In", `Out` = "Out", `Unknown` = "Unknown",
    `Moderately Differentiated` = "Moderately differentiated", 
    `Moderately-to-Poorly Differentiated` = "Moderately-to-Poorly differentiated",
    `Poorly Cohesive` = "poorly cohesive", `Poorly Differentiated` = "Poorly differentiated",
    `Well Differentiated` = "Well Differentiated", `Unknown` = "Unknown",
    `Stage 1` = "Stage 1", `Stage 2` = "Stage 2", `Stage 3` = "Stage 3", `Stage 4` = "Stage 4",
    `Alive` = "Alive (1)", `Alive (1+)` = "Alive (1+)", `Alive (2+)` = "Alive (2+)"
  )))

# Fix Length of Survival categories
data$Length.of.Survival..5.year. <- as.factor(data$Length.of.Survival..5.year.)
data$Length.of.Survival..5.year. <- as.character(data$Length.of.Survival..5.year.)

data$Length.of.Survival..5.year. <- recode(data$Length.of.Survival..5.year.,
  "<1" = "0", 
  "1" = "1", 
  "2" = "2", 
  "3" = "3", 
  "4" = "4", 
  "5+" = "5", 
  "Alive" = "5", 
  "Alive (1+)" = "1", 
  "Alive (2+)" = "2", 
  "Unknown" = NA_character_)

data$Length.of.Survival..5.year. <- factor(data$Length.of.Survival..5.year., 
                                            levels = c("0", "1", "2", "3", "4", "5"), 
                                            labels = c("0", "1", "2", "3", "4", "5"))

# Now, update the "White" race group to "Hispanic" if they also belong to a Hispanic origin group
data <- data %>%
  mutate(Race.1 = case_when(
    Race.1 == "White" & (Spanish.Hispanic.Origin %in% c("Mexican", "South or Central American", "Spanish")) ~ "Hispanic",
    TRUE ~ Race.1  # Keep the original race for other cases
  ))

```


```{r, echo=FALSE}
# Create a table of counts for Race.1 and Time.Period
#race_time_table <- table(data$Race.1, data$Time.Period)
#print(race_time_table)

```


```{r, echo=FALSE}
#start subsetting
# Assuming 'Year.of.Diagnosis' is a numeric variable
data$Time.Period <- ifelse(data$Year.of.Diagnosis >= 2010 & data$Year.of.Diagnosis <= 2016, "2010-2016", 
                           ifelse(data$Year.of.Diagnosis >= 2017 & data$Year.of.Diagnosis <= 2023, "2017-2023", NA))
```

```{r, echo=FALSE}
# Recode Hispanic status, including White individuals with Hispanic origin
data$Hispanic <- ifelse(data$Race.1 == "Hispanic", 1, 0)

# Recode African American status based on 'Race.1'
data$African.American <- ifelse(data$Race.1 == "Black or African American", 1, 0)

# Recode Born in US status based on 'Born.in.Vs.Out.of.US'
data$Born.in.Vs.Out.of.US <- trimws(data$Born.in.Vs.Out.of.US)

# Ensure there are no extra spaces
data$Born.in.Vs.Out.of.US <- trimws(data$Born.in.Vs.Out.of.US)

# Correct the assignment logic for Born.In.US
data$Born.In.US <- ifelse(data$Born.in.Vs.Out.of.US == "In", 1, 
                          ifelse(data$Born.in.Vs.Out.of.US == "Out", 0, NA))

```

```{r, echo=FALSE}
# Convert Race.1 to a factor
data$Race.1 <- as.factor(data$Race.1)

# Convert Time.Period to a factor
data$Time.Period <- as.factor(data$Time.Period)
```


```{r, echo=FALSE}
# Define the population data for both periods
total_population_2010_2016 <- c(2104289, 2127256, 2160074, 2194414, 2244323, 2285102, 2294844)
total_population_2017_2023 <- c(2311418, 2326927, 2324864, 2319550, 2360573, 2371448)

black_population_2010_2016 <- c(408232.07, 412687.66, 423374.50, 432299.56, 444375.95, 450165.09, 456673.96)
black_population_2017_2023 <- c(462283.60, 467712.33, 474272.26, 480146.85, 498080.90, 512232.77)

hispanic_population_2010_2016 <- c(862758.49, 880683.98, 896430.71, 912876.22, 938127.01, 959742.84, 973013.86)
hispanic_population_2017_2023 <- c(993909.74, 1007559.39, 1015965.57, 1029880.20, 1052815.56, 1045808.57)

# Calculate the averages for each period
avg_total_population_2010_2016 <- mean(total_population_2010_2016, na.rm = TRUE)
avg_total_population_2017_2023 <- mean(total_population_2017_2023, na.rm = TRUE)

avg_black_population_2010_2016 <- mean(black_population_2010_2016, na.rm = TRUE)
avg_black_population_2017_2023 <- mean(black_population_2017_2023, na.rm = TRUE)

avg_hispanic_population_2010_2016 <- mean(hispanic_population_2010_2016, na.rm = TRUE)
avg_hispanic_population_2017_2023 <- mean(hispanic_population_2017_2023, na.rm = TRUE)

# Create a simplified data frame with average values
estimated_at_risk_population_avg <- data.frame(
  Group = c("Total At-Risk Population", "Black Population", "Hispanic Population"),
  `2010-2016 Average` = c(avg_total_population_2010_2016, avg_black_population_2010_2016, avg_hispanic_population_2010_2016),
  `2017-2023 Average` = c(avg_total_population_2017_2023, avg_black_population_2017_2023, avg_hispanic_population_2017_2023)
)

# Print the data frame
#print(estimated_at_risk_population_avg)

# Title for the table
title <- "Estimated At-Risk Population Averages"

kable(estimated_at_risk_population_avg, caption = title)

```


```{r, echo=FALSE}
# Number of incidents (numerator) from the table for each race and period
incidents_black_2010_2016 <- 10
incidents_black_2017_2023 <- 4

incidents_hispanic_2010_2016 <- 44
incidents_hispanic_2017_2023 <- 52

incidents_total_2010_2016 <- sum(c(10, 44))  # Sum for Black + Hispanic (you can add others if needed)
incidents_total_2017_2023 <- sum(c(4, 52))   # Sum for Black + Hispanic (you can add others if needed)

# Calculate the incident rates for each group and period
incident_rate_black_2010_2016 <- (incidents_black_2010_2016 / avg_black_population_2010_2016) * 100000
incident_rate_black_2017_2023 <- (incidents_black_2017_2023 / avg_black_population_2017_2023) * 100000

incident_rate_hispanic_2010_2016 <- (incidents_hispanic_2010_2016 / avg_hispanic_population_2010_2016) * 100000
incident_rate_hispanic_2017_2023 <- (incidents_hispanic_2017_2023 / avg_hispanic_population_2017_2023) * 100000

incident_rate_total_2010_2016 <- (incidents_total_2010_2016 / avg_total_population_2010_2016) * 100000
incident_rate_total_2017_2023 <- (incidents_total_2017_2023 / avg_total_population_2017_2023) * 100000

# Create a data frame to store incident rates
incident_rates <- data.frame(
  Group = c("Black Population", "Hispanic Population", "Total Population"),
  `2010-2016 Incident Rate (per 100,000)` = c(incident_rate_black_2010_2016, incident_rate_hispanic_2010_2016, incident_rate_total_2010_2016),
  `2017-2023 Incident Rate (per 100,000)` = c(incident_rate_black_2017_2023, incident_rate_hispanic_2017_2023, incident_rate_total_2017_2023)
)

# Print the incident rates data frame
#print(incident_rates)

# Title for the table
incident_title <- "Incident Rates by Population Group"
#cat("\n", incident_title, "\n")

# Display the table in a formatted manner
kable(incident_rates, caption = incident_title)

```
```{r, echo=FALSE}
# Create the data frame for incident rates with cleaned column names
incident_rates <- data.frame(
  Group = c("Black Population", "Hispanic Population", "Total Population"),
  X2010_2016_Incident_Rate = c(2.311903, 4.794794, 2.452905),
  X2017_2023_Incident_Rate = c(0.8290932, 5.0765229, 2.3974690)
)

# Reshape the data for plotting
incident_rates_long <- gather(incident_rates, key = "Time_Period", value = "Incident_Rate", 
                              X2010_2016_Incident_Rate, X2017_2023_Incident_Rate)

ggplot(incident_rates_long, aes(x = Group, y = Incident_Rate, fill = Time_Period)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.4) +
  labs(
    title = "Incident Rates by Population Group",
    y = "Incident Rate (per 100,000)",
    x = "Population Group",
    fill = "Time Period"
  ) +
  scale_fill_manual(
    values = c("X2010_2016_Incident_Rate" = "skyblue", "X2017_2023_Incident_Rate" = "orange"),
    labels = c("2010-2016", "2017-2023")
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )

```

The bar graph displays the distribution of incidence rates for early-onset gastric cancer across different groups (Hispanic, African American, Total) and time periods (2010-2016 and 2017-2023).
The color coding differentiates between the groups, and the facet grids separate the data by time period.
From this graph, you can compare how the incidence rate of early-onset gastric cancer varies across different groups and time periods, looking for any notable differences or trends.

```{r, echo=FALSE}
# Reshape the population data for merging
estimated_at_risk_population_avg_long <- gather(estimated_at_risk_population_avg,
                                                 key = "Time_Period", value = "Population_Avg", 
                                                 X2010.2016.Average, X2017.2023.Average)

# Standardize the 'Group' names in 'estimated_at_risk_population_avg_long'
estimated_at_risk_population_avg_long$Group <- gsub("Total At-Risk Population", "Total Population", 
                                                     estimated_at_risk_population_avg_long$Group)

# Standardize the 'Time_Period' names in both data frames
estimated_at_risk_population_avg_long$Time_Period <- gsub("X2010.2016.Average", "X2010_2016_Incident_Rate", 
                                                           estimated_at_risk_population_avg_long$Time_Period)
estimated_at_risk_population_avg_long$Time_Period <- gsub("X2017.2023.Average", "X2017_2023_Incident_Rate", 
                                                           estimated_at_risk_population_avg_long$Time_Period)
# Now merge the two data frames
bubble_data <- merge(estimated_at_risk_population_avg_long, incident_rates_long, 
                     by = c("Group", "Time_Period"))

# Check the merged data (optional but recommended for debugging)
#head(bubble_data)

# Create a nicer label for Time_Period
time_period_labels <- c(
  "X2010_2016_Incident_Rate" = "2010-2016",
  "X2017_2023_Incident_Rate" = "2017-2023"
)

bubble_plot <- ggplot(bubble_data, aes(x = Population_Avg, y = Incident_Rate, 
                                       size = Population_Avg, color = Group)) +
  geom_point(alpha = 0.7) +
  facet_wrap(~ Time_Period, scales = "free_y", labeller = as_labeller(time_period_labels)) +  # Apply nicer labels
  scale_size_continuous(range = c(5, 15)) +  # Adjust bubble size range
  labs(title = "Bubble Plot of Population Averages and Incident Rates",
       x = "Population Average",
       y = "Incident Rate",
       size = "Population Size",
       color = "Group") +
  coord_cartesian(ylim = c(-1, 7)) +  # Using coord_cartesian to avoid clipping
  xlim(425000, 3025000) +  # Adjust x limits
  theme_minimal() +
  theme(
    legend.position = "top",    # Place the legend on top
    legend.box = "vertical",    # Stack legend items vertically
    legend.margin = margin(0, 0, 10, 0),  # Adjust margin between legend and plot
    legend.key.size = unit(0.5, "cm"),    # Reduce the size of the legend keys
    plot.margin = margin(10, 10, 50, 10),  # Adjust bottom margin
    legend.text = element_text(size = 8),  # Reduce font size of legend text
    legend.title = element_text(size = 10)  # Reduce font size of legend title
  )

# Display the plot
print(bubble_plot)

```
This bubble plot visualizes the relationship between population averages and incident rates for different groups (Black, Hispanic, and Total Population) across two time periods: 2010-2016 and 2017-2023. The size of each bubble corresponds to the population average, while the y-axis represents the incident rate, with the data split by time period to highlight trends over time.


Descriptive Statistics

```{r, echo=FALSE}
# One-way ANOVA for comparing incident rates across groups and time periods
anova_results <- aov(Incident_Rate ~ Group * Time_Period, data = incident_rates_long)

# Summarize the ANOVA results and print as a table
anova_summary <- summary(anova_results)
anova_summary_table <- as.data.frame(anova_summary[[1]])  # Extracting ANOVA summary and converting to data frame
kable(anova_summary_table, caption = "ANOVA Summary for Incident Rates across Groups and Time Periods")

# Calculate percentage change for each group in incident_rates dataframe
incident_rates_wide <- incident_rates_long %>%
  spread(Time_Period, Incident_Rate) %>%
  mutate(
    Percent_Change = (`X2017_2023_Incident_Rate` - `X2010_2016_Incident_Rate`) /
                     `X2010_2016_Incident_Rate` * 100
  )

# View the results and print as a nice table using kable
kable(incident_rates_wide, caption = "Incident Rates and Percent Change by Group")

```
Descriptive Statistics Overview:
Summary of ANOVA Results:
The ANOVA analysis evaluated the effects of Group, Time_Period, and their interaction on the incident rates. The results indicated that neither the Group factor nor the Time_Period factor had a statistically significant impact on incident rates. Additionally, there was no significant interaction between these factors (all p-values greater than 0.05). This suggests that incident rates did not vary meaningfully by group or time period across the populations studied.

Summary of Percent Change in Incident Rates:
The percent change in incident rates between the periods 2010-2016 and 2017-2023 revealed different trends across groups:

Black Population experienced a significant decrease in incident rates, with a 64.14% reduction.
Hispanic Population showed a modest 5.88% increase in incident rates.
Total Population saw a slight 2.26% decrease in incident rates. These changes suggest varying trends, but they were not statistically significant in the ANOVA analysis.

Sources of Error:
However, the analysis is limited by the small sample size which prevents any meaningful computation of variability or statistical significance. Consequently, while we can describe the trends in incidence rates, further data collection is required to assess the true variability and test for statistical significance. This would provide a deeper and more reliable understanding of the trends observed across different demographic groups. 

A source of error may be in the average population estimation. Population data for the at-risk age range was not widely available, so the age range 20-54 was used to calculate the estimates, as that was the closest age range available. Population data for 2020 was not available, making the average for the 2017-2023 age range less strong of an average. 



